---
title: "Figure_brain"
author: "Kimberly Olney"
date: "06/15/2022"
output: html_document
---

Main figure showing brain results

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
```

May to unload all packages 
lapply(paste('package:',names(sessionInfo()$otherPkgs),sep=""),detach,character.only=TRUE,unload=TRUE)
```{r packages}
library(gprofiler2)
library(DESeq2)
library(enrichplot)
library(DOSE) 
library(plyr)
library(ggplot2)
require(gridExtra)
library(forcats)
library(UpSetR)
library(ComplexUpset)
library(ggplot2movies)

library(gtools)
library(ggnewscale)
library(ggrepel)

library(patchwork)
require(gridExtra)
library(ggpubr)
library(scales)
library("wiggleplotr")
library("dplyr")
library("GenomicRanges")
library("GenomicFeatures")
library("biomaRt")
library(tidyverse)
library(sleuth)
```

# User defined variables
```{r set_variables}
tissue <- "Brain"
control <- "Saline"
treatment <- "LPS"
control_color <- "gray29"
treatment_color <- "purple"
pathToRef <- c("/research/labs/neurology/fryer/projects/references/pig/")
```
# Save functions
These functions with help simultaneously save plots as a png, pdf, and tiff 
file.
```{r}
saveToPDF <- function(...) {
    d = dev.copy(pdf,...)
    dev.off(d)
}

saveToPNG <- function(...) {
    d = dev.copy(png,...)
    dev.off(d)
}
```

# Binary list shared gene function
```{r}
fromList <- function (input) {
  # Same as original fromList()...
  elements <- unique(unlist(input))
  data <- unlist(lapply(input, function(x) {
      x <- as.vector(match(elements, x))
      }))
  data[is.na(data)] <- as.integer(0)
  data[data != 0] <- as.integer(1)
  data <- data.frame(matrix(data, ncol = length(input), byrow = F))
  data <- data[which(rowSums(data) != 0), ]
  names(data) <- names(input)
  # ... Except now it conserves your original value names!
  row.names(data) <- elements
  return(data)
  }
```

# Gene level 
### Format dataframe
```{r}
tool = c("star")

treatment_vs_control <- read.table(
  paste0(
    "../../../results/",
    tool,
    "/DEGs/",
    treatment,
    "_",
    tissue,
    "_gene_DEGs_FDRq1.00.txt",
    sep = ""
  ),
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)

color_values <- vector()
max <- nrow(treatment_vs_control)

for (i in 1:max) {
  if (treatment_vs_control$adj.P.Val[i] < 0.05) {
    if (treatment_vs_control$logFC[i] > 0) {
      color_values <-
        c(color_values, 1) # 1 when logFC > 0 and FDRq < 0.05
    }
    else if (treatment_vs_control$logFC[i] < 0) {
      color_values <-
        c(color_values, 2) # 2 when logFC < 0 and FDRq < 0.05
    }
  }
  else{
    color_values <- c(color_values, 3) # 3 when FDRq >= 0.05
  }
}

treatment_vs_control$color_p0.05 <- factor(color_values)
up <- treatment_vs_control[treatment_vs_control$color_p0.05 == 1, ]
up10 <- up[1:5, ]

upFoldChange <-
  treatment_vs_control[treatment_vs_control$logFC > 4, ]
upFoldChange <- upFoldChange[5:10, ]

down <- treatment_vs_control[treatment_vs_control$color_p0.05 == 2, ]
down <- subset(down, down$logFC < -1.5)
down10 <- down[1:5, ]

downFoldChange <-
  treatment_vs_control[treatment_vs_control$logFC < -3 &
                         treatment_vs_control$adj.P.Val < 0.05, ]
downFoldChange <- downFoldChange[5:10, ]

goi <-
  treatment_vs_control[treatment_vs_control$gene_name == "AP4M1", ]
goi2 <-
  treatment_vs_control[treatment_vs_control$gene_name == "SOCS3", ]

hadjpval <- (-log10(max(treatment_vs_control$P.Value[treatment_vs_control$adj.P.Val < 0.05],
                        na.rm = TRUE)))
```
### Volcano plot
```{r}
gene_volcano <-
  ggplot(data = treatment_vs_control,
         aes(
           x = logFC,
           # x-axis is logFC
           y = -log10(P.Value),
           # y-axis will be -log10 of P.Value
           color = color_p0.05
         )) +  # color is based on factored color column
  geom_point(alpha = 0.5, size = 1.7) +
  theme_bw() +  # set color theme
  theme(legend.position = "none") +  # no legend
  scale_color_manual(values = c("red", "blue", "grey")) +  # set factor colors
  labs(x = expression(log[2](FC)),
       y = expression(-log[10] ~ "(" ~ italic("p") ~ "-value)")) +
       theme(axis.title.x = element_text(size = 10),
             axis.text.x = element_text(size = 10)) +
         theme(axis.title.y = element_text(size = 10),
               axis.text.y = element_text(size = 10)) +
         geom_hline(yintercept = hadjpval,
                    #  horizontal line
                    colour = "#000000",
                    linetype = "dashed") +
        ggtitle(paste0(tissue,"\nFDRq < 0.05")) +
         theme(plot.title = element_text(size = 10)) +
         geom_text_repel(
           data = up10,
           aes(
             x = logFC,
             y = -log10(P.Value),
             label = gene_name
           ),
           color = "maroon",
           fontface = "italic",
           size = 2.7,
           min.segment.length = unit(0, 'lines'),
           nudge_x = .25,
           max.overlaps = getOption("ggrepel.max.overlaps", default = 30)
         ) +
         geom_text_repel(
           data = upFoldChange,
           aes(
             x = logFC,
             y = -log10(P.Value),
             label = gene_name
           ),
           color = "maroon",
           fontface = "italic",
           size = 2.7,
           min.segment.length = unit(0, 'lines'),
           nudge_x = .25,
           max.overlaps = getOption("ggrepel.max.overlaps", default = 30)
         ) +
         geom_text_repel(
           data = down10,
           aes(
             x = logFC,
             y = -log10(P.Value),
             label = gene_name
           ),
           color = "navyblue",
           fontface = "italic",
           size = 2.7,
           min.segment.length = unit(0, 'lines'),
           nudge_x = -1,
           nudge_y = .25,
           max.overlaps = getOption("ggrepel.max.overlaps", default = 15)
         )  +
         geom_text_repel(
           data = downFoldChange,
           aes(
             x = logFC,
             y = -log10(P.Value),
             label = gene_name
           ),
           color = "navyblue",
           fontface = "italic",
           size = 2.7,
           min.segment.length = unit(0, 'lines'),
           nudge_x = -.5,
           max.overlaps = getOption("ggrepel.max.overlaps", default = 15)
         ) +
         geom_text_repel(
           data = goi,
           aes(
             x = logFC,
             y = -log10(P.Value),
             label = gene_name
           ),
           color = "black",
           fontface = "italic",
           size = 2.7,
           min.segment.length = unit(0, 'lines'),
           max.overlaps = getOption("ggrepel.max.overlaps", default = 5)
         ) +
         geom_text_repel(
           data = goi2,
           aes(
             x = logFC,
             y = -log10(P.Value),
             label = gene_name
           ),
           color = "black",
           fontface = "italic",
           size = 2.7,
           min.segment.length = unit(0, 'lines'),
           max.overlaps = getOption("ggrepel.max.overlaps", default = 5)
         ) +
         scale_y_continuous(breaks = seq(0, 12, by = 1), limits = c(0, 12)) +
         scale_x_continuous(breaks = seq(-8, 8, by = 2), limits = c(-8, 8))
gene_volcano
```
remove files no longer needed
```{r}
remove(
  down,
  down10,
  downFoldChange,
  goi,
  goi2,
  up,
  up10,
  upFoldChange,
  hadjpval,
  i,
  max,
  color_values
)
```

# Metascape Enrichment 
### Format the dataframes
```{r}
# read in enrichment analysis results
up_enrich_results <-
  read.delim(
    paste0(
      "../../../results/",
      tool,
      "/metascape/",
      tolower(tissue),
      "_up.txt"
    ),
    sep = "\t",
    header = TRUE
  )

down_enrich_results <-
  read.delim(
    paste0(
      "../../../results/",
      tool,
      "/metascape/",
      tolower(tissue),
      "_down.txt"
    ),
    sep = "\t",
    header = TRUE
  )

# select the GO term IDs we want to show in the plot
GO_ID_up <-
  c(
    "1_Summary",
    "2_Summary",
    "3_Summary",
    "4_Summary",
    "5_Summary",
    "6_Summary",
    "7_Summary",
    "8_Summary",
    "9_Summary", 
    "10_Summary",
    "11_Summary",
    "12_Summary",
    "13_Summary",
    "14_Summary",
  #  "15_Summary",
  #  "16_Summary",
    #"17_Summary",
   # "18_Summary",
   # "19_Summary", 
    "20_Summary"
  )
GO_ID_down <- c(
    "1_Summary",
    "2_Summary",
    "3_Summary",
    "4_Summary",
    "5_Summary",
    "6_Summary",
    "7_Summary",
    "8_Summary",
    "9_Summary", 
    "10_Summary",
    "11_Summary",
    "12_Summary",
    "13_Summary",
    "14_Summary",
    "15_Summary"
  #  "16_Summary",
  #  "17_Summary",
  #  "18_Summary",
  #  "19_Summary", 
   # "20_Summary"
)

up_enrich_results_subset <-
  up_enrich_results[up_enrich_results$GroupID %in% GO_ID_up, ]
up_enrich_results_subset$Cluster <- c("up-regulated")
up_enrich_results_subset$Description <-
  factor(up_enrich_results_subset$Description,
         levels = up_enrich_results_subset$Description)
up_enrich_results_subset$Description <-
  fct_rev(up_enrich_results_subset$Description)

down_enrich_results_subset <-
  down_enrich_results[down_enrich_results$GroupID %in% GO_ID_down, ]
down_enrich_results_subset$Cluster <- c("down-regulated")
down_enrich_results_subset$Description <-
  factor(down_enrich_results_subset$Description,
         levels = down_enrich_results_subset$Description)
down_enrich_results_subset$Description <-
  fct_rev(down_enrich_results_subset$Description)

# get the number of genes in each summary
up_gene_count <-
  strsplit(as.character(up_enrich_results_subset$InTerm_InList),
           "/",
           fixed = T)
up_gene_count_df <-
  data.frame(matrix(
    unlist(up_gene_count),
    nrow = length(up_gene_count),
    byrow = TRUE
  ), stringsAsFactors = FALSE)
up_enrich_results_subset$InTerm <- as.numeric(up_gene_count_df$X1)
up_enrich_results_subset$InList <- as.numeric(up_gene_count_df$X2)

down_gene_count <-
  strsplit(as.character(down_enrich_results_subset$InTerm_InList),
           "/",
           fixed = T)
down_gene_count_df <-
  data.frame(matrix(
    unlist(down_gene_count),
    nrow = length(down_gene_count),
    byrow = TRUE
  ), stringsAsFactors = FALSE)
down_enrich_results_subset$InTerm <- as.numeric(down_gene_count_df$X1)
down_enrich_results_subset$InList <- as.numeric(down_gene_count_df$X2)

# combine together
up_and_down_enrich_results_subset <- rbind(up_enrich_results_subset, down_enrich_results_subset)

up_and_down_enrich_results_subset$Description <-
      gsub(
        "regulation of plasma membrane bounded cell projection organization",
        "reg of plasma membrane/cell projection organization",
        up_and_down_enrich_results_subset$Description
      )
up_and_down_enrich_results_subset$Description <-
  factor(up_and_down_enrich_results_subset$Description, levels = unique(up_and_down_enrich_results_subset$Description)) 

up_and_down_enrich_results_subset$Description <- fct_rev(up_and_down_enrich_results_subset$Description)
```

remove files 
```{r}
remove(
  down_enrich_results,
  down_gene_count,
  down_gene_count_df,
  up_enrich_results,
  up_gene_count,
  up_gene_count_df
)
```

### Enrichment plot showing log10 P-value
#### down and up seperately
```{r}
down_enrich_plot <-
  ggplot(data = down_enrich_results_subset, aes(x = InTerm, y = Description)) +
  ggplot2::facet_grid(~ Cluster) +
  geom_bar(stat = "identity", aes(fill = LogP), width = .65) +
  theme(axis.title.x = element_text(size = 10),
        axis.text.x = element_text(size = 10)) +
  theme(axis.title.y = element_text(size = 10),
        axis.text.y = element_text(size = 10)) +
  theme_bw() +
  scale_y_discrete(expand = c(0, 0)) +
  labs(x = "Gene count", y = "") +
  #  xlim(0,60) +
  guides(fill = guide_legend(
    size = 1,
    title = expression(log[10] ~ "(" ~ italic("p") ~ "-value)")
  )) +
  scale_fill_gradientn(
    colours = c("darkmagenta", "mediumorchid3", "grey"),
    guide = "legend",
    limits = c(-45,-1),
  ) +
  theme(legend.text = element_text(size = 10),
        legend.title = element_text(size = 10))
down_enrich_plot

up_enrich_plot <-
  ggplot(data = up_enrich_results_subset, aes(x = InTerm, y = Description)) +
  ggplot2::facet_grid(~ Cluster) +
  geom_bar(stat = "identity", aes(fill = LogP), width = .65) +
  theme(axis.title.x = element_text(size = 10),
        axis.text.x = element_text(size = 10)) +
  theme(axis.title.y = element_text(size = 10),
        axis.text.y = element_text(size = 10)) +
  theme(legend.text = element_text(size = 6)) +
  theme_bw() +
  scale_y_discrete(expand = c(0, 0)) +
  labs(x = "Gene count", y = "") +
  guides(fill = guide_legend(title = expression(log[10] ~ "(" ~ italic("p") ~ "-value)"))) +
  scale_fill_gradientn(
    colours = c("darkmagenta", "mediumorchid3", "grey"),
    guide = "legend",
    limits = c(-45,-1)
  ) +
  theme(legend.text = element_text(size = 10),
        legend.title = element_text(size = 10))
up_enrich_plot
```

#### down and up together
```{r}
up_and_down_enrich_plot <-
  ggplot(data = up_and_down_enrich_results_subset, aes(x = InTerm, y = Description)) +
  ggplot2::facet_grid(~ Cluster, scales = "free") +
  geom_bar(stat = "identity", aes(fill = LogP), width = .65) +
  theme(axis.title.x = element_text(size = 10),
        axis.text.x = element_text(size = 10)) +
  theme(axis.title.y = element_text(size = 10),
        axis.text.y = element_text(size = 9)) +
  theme(legend.text = element_text(size = 6)) +
  theme_bw() +
  ggtitle(paste0(tissue," enrichment summaries")) +
  theme(plot.title = element_text(size = 10)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(x = "Gene count", y = NULL) +
  guides(fill = guide_legend(title = expression(log[10] ~ "(" ~ italic("p") ~ "-value)"))) +
  scale_fill_gradientn(
    colours = c("darkmagenta", "mediumorchid3", "grey"),
    guide = "legend",
    limits = c(-45,-1)
  ) +
  theme(legend.text = element_text(size = 8.5),
        legend.title = element_text(size = 8.5),
        legend.margin=margin(0,0.5,0,0),
        legend.box.margin=margin(-10,-2,-10,-7.5)) +
  theme(plot.title = element_text(hjust = -3.5, vjust=0))
up_and_down_enrich_plot
```

```{r}
remove(up_enrich_plot,
       down_enrich_plot)
```
## Combine volcano and enrichment plot 
```{r}
row1 <- ggarrange(
  gene_volcano,
  up_and_down_enrich_plot,
  ncol = 2,
  labels = c("a)", "b)"),
  widths = c(3.5, 6.5), 
  font.label = list(size = 11)
  )
row1

path <- paste0("../../../results/manuscript_figures/Figure_",
         tolower(tissue),
         "row1_test")
saveToPDF(paste0(path, ".pdf"), width = 10, height = 4.5)
```
# CPM data
```{r}
# read in counts per million
cpm <- read.delim(paste0("../../../results/", tool, "/CPM/", treatment, "_",tolower(tissue),
               "_gene_CPM_expression.txt"))
# then log2 transform 
lcpm <- edgeR::cpm(cpm, log = TRUE)
# save ouput
write.table(lcpm, paste0("../../../results/", tool, "/CPM/", treatment, "_",tolower(tissue),
               "_gene_log2CPM_expression.txt"), quote = FALSE, sep = "\t")
```
### Fold change information 
```{r}
up_regulated <- subset(
  treatment_vs_control,
  treatment_vs_control$adj.P.Val <= 0.05 &
    treatment_vs_control$logFC > 0
)
down_regulated <- subset(
  treatment_vs_control,
  treatment_vs_control$adj.P.Val <= 0.05 &
    treatment_vs_control$logFC <= 0
)

# sort by log2FC and adjusted p-val
up_regulated_sort <-
  up_regulated[order(-up_regulated$logFC, up_regulated$adj.P.Val), ]
down_regulated_sort <-
  down_regulated[order(down_regulated$logFC, down_regulated$adj.P.Val), ]

# get the lcpm information for those genes of interest
up_regulated_lcpm <- subset(lcpm, row.names(lcpm) %in%
                              up_regulated_sort$gene_name)
down_regulated_lcpm <- subset(lcpm, row.names(lcpm) %in%
                                down_regulated_sort$gene_name)

# reform the data
up_regulated_sort_df <- up_regulated_lcpm %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(-c(gene), names_to = "samples", values_to = "counts") %>%
  mutate(samples = fct_relevel(samples, colnames(up_regulated_lcpm)))

down_regulated_sort_df <- down_regulated_lcpm %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(-c(gene), names_to = "samples", values_to = "counts") %>%
  mutate(samples = fct_relevel(samples, colnames(down_regulated_lcpm)))

# lock in gene order which is sorted by adjusted p-value
up_regulated_sort_df$gene <-
  factor(up_regulated_sort_df$gene,
         levels = unique(up_regulated_sort_df$gene))
up_regulated_sort_df$gene <-
  fct_rev(up_regulated_sort_df$gene)
down_regulated_sort_df$gene <-
  factor(down_regulated_sort_df$gene,
         levels = unique(down_regulated_sort_df$gene))
down_regulated_sort_df$gene <-
  fct_rev(down_regulated_sort_df$gene)

DEG_df <-
  rbind(up_regulated_sort_df, down_regulated_sort_df)

DEG_df$Condition <-
  ifelse(grepl("LPS", DEG_df$samples),
         "LPS",
         "Control")

DEG_df$samples <- factor(
  DEG_df$samples,
  levels = c(
    "X3.Control",
    "X4.Control",
    "X6.Control",
    "X7.Control",
    "X10.Control",
    "X11.Control",
    "X2.LPS",
    "X8.LPS",
    "X12.LPS",
    "X14.LPS"
  )
)
DEG_sort <- rbind(up_regulated_sort, down_regulated_sort)
names(DEG_sort)[names(DEG_sort) == "gene_name"] <- "gene"

```
remove files
```{r}
remove(
  cpm,
  lcpm,
  down_regulated,
  down_regulated_lcpm,
  down_regulated_sort,
  down_regulated_sort_df,
  up_regulated,
  up_regulated_lcpm,
  up_regulated_sort,
  up_regulated_sort_df
)
```

### Genes in enrichment plot all
## up 
vasculature development	[20]
Cytokine Signaling in Immune system	[1]
TNF signaling pathway	[5]
NF-kappa B signaling pathway	[9]
positive regulation of cell death [13]
```{r}
up_enrich_df <-
  up_enrich_results_subset[, c("GroupID", "Term", "Description", "Symbols")]
up_enrich_df_melt <- reshape2::melt(up_enrich_df)
up_gene_char <- str_split(up_enrich_df_melt$Symbols, ",")

up_gene_pathway <-
  data.frame(gene = unlist(up_gene_char),
             value = rep(up_enrich_df$Description, lengths(up_gene_char)))
up_pathways_levels <- levels(fct_rev(up_gene_pathway$value))

up_input <- list(
  "vasculature development" = up_gene_char[[15]],
  "Cytokine Signaling in Immune system" = up_gene_char[[1]],
  "NF signaling pathway" = up_gene_char[[5]],
  "NF-kappa B signaling pathway" = up_gene_char[[9]], 
  "positive regulation of cell death" = up_gene_char[[11]]
)
up_data <- fromList(up_input)
up_data$gene <- row.names(up_data)
up_data_melt <- reshape2::melt(up_data)


# get the fold change value for those genes
up_df <- merge(up_data_melt, DEG_sort,
                       by = "gene")
# order by fold change 
up_df <- up_df[order(-up_df$logFC, -up_df$adj.P.Val),]

# subset by GO termes
GO_1 <-
  subset(up_df,
         up_df$variable == "vasculature development"
         & up_df$value == 1)
GO_1_top <- head(GO_1, 15)

GO_2 <-
  subset(up_df,
         up_df$variable == "Cytokine Signaling in Immune system"
         & up_df$value == 1)
GO_2_top <- head(GO_2, 15)

GO_3 <-
  subset(up_df,
         up_df$variable == "NF signaling pathway"
         & up_df$value == 1)
GO_3_top <- head(GO_3, 15)

GO_4 <-
  subset(up_df,
         up_df$variable == "NF-kappa B signaling pathway"
         & up_df$value == 1)
GO_4_top <- head(GO_4, 15)

GO_5 <-
  subset(up_df,
         up_df$variable == "positive regulation of cell death"
         & up_df$value == 1)
GO_5_top <- head(GO_5, 15)
```

#### plot lcpm GO heatmaps
##### up 1
```{r}
# lock in gene order which is sorted by adjusted p-value
GO_1_DEG <- subset(DEG_df, DEG_df$gene %in% GO_1_top$gene)
GO_1_DEG$gene <- factor(GO_1_DEG$gene,
                          levels = unique(GO_1_top$gene))
GO_1_DEG$gene <- fct_rev(GO_1_DEG$gene)
# heatmap 
up_GO_1_lcpm_heat <- ggplot(data = GO_1_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    name = expression(log[2](CPM))
  ) +
  scale_x_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(0.2, 0.5, 0.2, 0.2, "cm"), 
    panel.spacing = unit(0,'lines'), 
    plot.title = element_text(size = 9, vjust = -1, hjust = 0.725),
    axis.title.x=element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks.x=element_blank(), 
    axis.title.y = element_blank(), 
    axis.text.y = element_text(size = 8, margin = margin(r = -2)),
    axis.ticks.y=element_blank()) +
    ggtitle(paste0(up_pathways_levels[[15]])) 
up_GO_1_lcpm_heat
```
##### up 2
```{r}
# lock in gene order which is sorted by adjusted p-value
GO_2_DEG <- subset(DEG_df, DEG_df$gene %in% GO_2_top$gene)
GO_2_DEG$gene <- factor(GO_2_DEG$gene,
                          levels = unique(GO_2_top$gene))
GO_2_DEG$gene <- fct_rev(GO_2_DEG$gene)

# heatmap 
up_GO_2_lcpm_heat <- ggplot(data = GO_2_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    name = expression(log[2](CPM))
  ) +
  scale_x_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(0.2, 0.5, 0.2, 0.2, "cm"), 
    panel.spacing = unit(0,'lines'), 
    plot.title = element_text(size = 9, vjust = -1, hjust = .6),
    axis.title.x=element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks.x=element_blank(), 
    axis.title.y = element_blank(), 
    axis.text.y = element_text(size = 8, margin = margin(r = -2)),
    axis.ticks.y=element_blank()) +
    ggtitle(paste0(up_pathways_levels[[2]]))
up_GO_2_lcpm_heat
```
##### up 3
```{r}
# lock in gene order which is sorted by adjusted p-value
GO_3_DEG <- subset(DEG_df, DEG_df$gene %in% GO_3_top$gene)
GO_3_DEG$gene <- factor(GO_3_DEG$gene,
                          levels = unique(GO_3_top$gene))
GO_3_DEG$gene <- fct_rev(GO_3_DEG$gene)

# heatmap 
up_GO_3_lcpm_heat <- ggplot(data = GO_3_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    name = expression(log[2](CPM))
  ) +
  scale_x_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(0.2, 0.5, 0.2, 0.2, "cm"), 
    panel.spacing = unit(0,'lines'), 
    plot.title = element_text(size = 9, vjust = -1, hjust = .6),
    axis.title.x=element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks.x=element_blank(), 
    axis.title.y = element_blank(), 
    axis.text.y = element_text(size = 8, margin = margin(r = -2)),
    axis.ticks.y=element_blank()) +
    ggtitle(paste0(up_pathways_levels[[5]]))
up_GO_3_lcpm_heat
```
##### up 4
```{r}
# lock in gene order which is sorted by adjusted p-value
GO_4_DEG <- subset(DEG_df, DEG_df$gene %in% GO_4_top$gene)
GO_4_DEG$gene <- factor(GO_4_DEG$gene,
                          levels = unique(GO_4_top$gene))
GO_4_DEG$gene <- fct_rev(GO_4_DEG$gene)

# heatmap 
up_GO_4_lcpm_heat <- ggplot(data = GO_4_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    name = expression(log[2](CPM))
  ) +
  scale_x_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(0.2, 0.5, 0.2, 0.2, "cm"), 
    panel.spacing = unit(0,'lines'), 
    plot.title = element_text(size = 9, vjust = -1, hjust = .6),
    axis.title.x=element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks.x=element_blank(), 
    axis.title.y = element_blank(), 
    axis.text.y = element_text(size = 8, margin = margin(r = -2)),
    axis.ticks.y=element_blank()) +
    ggtitle(paste0(up_pathways_levels[[9]]))
up_GO_4_lcpm_heat
```

##### up 5
```{r}
# lock in gene order which is sorted by adjusted p-value
GO_5_DEG <- subset(DEG_df, DEG_df$gene %in% GO_5_top$gene)
GO_5_DEG$gene <- factor(GO_5_DEG$gene,
                          levels = unique(GO_5_top$gene))
GO_5_DEG$gene <- fct_rev(GO_5_DEG$gene)

# heatmap 
up_GO_5_lcpm_heat <- ggplot(data = GO_5_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    name = expression(log[2](CPM))
  ) +
  scale_x_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(0.2, 0.5, 0.2, 0.2, "cm"), 
    panel.spacing = unit(0,'lines'), 
    plot.title = element_text(size = 9, vjust = -1, hjust=.6),
    axis.title.x=element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks.x=element_blank(), 
    axis.title.y = element_blank(), 
    axis.text.y = element_text(size = 8, margin = margin(r = -2)),
    axis.ticks.y=element_blank()) +
    ggtitle(paste0(up_pathways_levels[[11]]))
up_GO_5_lcpm_heat
```

clean up
```{r, eval = FALSE}
remove(
  up_data,
  up_data_melt,
  up_df,
  up_enrich_df,
  up_enrich_df_melt,
  up_enrich_results_subset,
  up_gene_char,
  up_gene_pathway,
  up_input,
  GO_1,
  GO_1_DEG,
  GO_1_top,
  GO_2,
  GO_2_DEG,
  GO_2_top,
  GO_3,
  GO_3_DEG,
  GO_3_top,
  GO_4,
  GO_4_DEG,
  GO_4_top,
  GO_5, 
  GO_5_DEG,
  GO_5_top
)
```
## down
vasculature development [1]
immune system development	[2]
RAC1 GTPase cycle	tight junction assembly	[4]
tight junction assembly [14]
Hemostasis [9]
or	TYROBP causal network in microglia
```{r}
down_enrich_df <-
  down_enrich_results_subset[, c("GroupID", "Term", "Description", "Symbols")]
down_enrich_df_melt <- reshape2::melt(down_enrich_df)
down_gene_char <- str_split(down_enrich_df_melt$Symbols, ",")

down_gene_pathway <-
  data.frame(gene = unlist(down_gene_char),
             value = rep(down_enrich_df$Description, lengths(down_gene_char)))
down_pathways_levels <- levels(fct_rev(down_gene_pathway$value))

down_input <- list(
  "vasculature development" = down_gene_char[[1]],
  "immune system development" = down_gene_char[[2]],
  "RAC1 GTPase cycle" = down_gene_char[[4]], 
  "Hemostasis" = down_gene_char[[9]],
  "tight junction assembly" = down_gene_char[[14]]
)
down_data <- fromList(down_input)
down_data$gene <- row.names(down_data)
down_data_melt <- reshape2::melt(down_data)


# get the fold change value for those genes
down_df <- merge(down_data_melt, DEG_sort,
                       by = "gene")
# order by fold change 
down_df <- down_df[order(down_df$logFC, -down_df$adj.P.Val),]

# subset by GO termes
GO_1 <-
  subset(down_df,
         down_df$variable == "vasculature development"
         & down_df$value == 1)
GO_1_top <- head(GO_1, 15)

GO_2 <-
  subset(down_df,
         down_df$variable == "immune system development"
         & down_df$value == 1)
GO_2_top <- head(GO_2, 15)

GO_3 <-
  subset(down_df,
         down_df$variable == "RAC1 GTPase cycle"
         & down_df$value == 1)
GO_3_top <- head(GO_3, 15)

GO_4 <-
  subset(down_df,
         down_df$variable == "Hemostasis"
         & down_df$value == 1)
GO_4_top <- head(GO_4, 15)

GO_5 <-
  subset(down_df,
         down_df$variable == "tight junction assembly"
         & down_df$value == 1)
GO_5_top <- head(GO_5, 15)
```
#### plot lcpm GO heatmaps
##### down 1
```{r}
# lock in gene order which is sorted by adjusted p-value
GO_1_DEG <- subset(DEG_df, DEG_df$gene %in% GO_1_top$gene)
GO_1_DEG$gene <- factor(GO_1_DEG$gene,
                          levels = unique(GO_1_top$gene))
GO_1_DEG$gene <- fct_rev(GO_1_DEG$gene)

# heatmap 
down_GO_1_lcpm_heat <- ggplot(data = GO_1_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    name = expression(log[2](CPM))
  ) +
  scale_x_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(0.2, 0.5, 0.2, 0.2, "cm"), 
    panel.spacing = unit(0,'lines'), 
    plot.title = element_text(size = 9, vjust = -1, hjust = .6),
    axis.title.x=element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks.x=element_blank(), 
    axis.title.y = element_blank(), 
    axis.text.y = element_text(size = 8, margin = margin(r = -2)),
    axis.ticks.y=element_blank()) +
    ggtitle(paste0(down_pathways_levels[[1]])) 
down_GO_1_lcpm_heat
```
##### down 2
```{r}
# lock in gene order which is sorted by adjusted p-value
GO_2_DEG <- subset(DEG_df, DEG_df$gene %in% GO_2_top$gene)
GO_2_DEG$gene <- factor(GO_2_DEG$gene,
                          levels = unique(GO_2_top$gene))
GO_2_DEG$gene <- fct_rev(GO_2_DEG$gene)

# heatmap 
down_GO_2_lcpm_heat <- ggplot(data = GO_2_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    name = expression(log[2](CPM))
  ) +
  scale_x_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(0.2, 0.5, 0.2, 0.2, "cm"), 
    panel.spacing = unit(0,'lines'), 
    plot.title = element_text(size = 9, vjust = -1, hjust = .6),
    axis.title.x=element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks.x=element_blank(), 
    axis.title.y = element_blank(), 
    axis.text.y = element_text(size = 8, margin = margin(r = -2)),
    axis.ticks.y=element_blank()) +
    ggtitle(paste0(down_pathways_levels[[2]]))
down_GO_2_lcpm_heat
```
##### down 3
```{r}
# lock in gene order which is sorted by adjusted p-value
GO_3_DEG <- subset(DEG_df, DEG_df$gene %in% GO_3_top$gene)
GO_3_DEG$gene <- factor(GO_3_DEG$gene,
                          levels = unique(GO_3_top$gene))
GO_3_DEG$gene <- fct_rev(GO_3_DEG$gene)

# heatmap 
down_GO_3_lcpm_heat <- ggplot(data = GO_3_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    name = expression(log[2](CPM))
  ) +
  scale_x_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(0.2, 0.5, 0.2, 0.2, "cm"), 
    panel.spacing = unit(0,'lines'), 
    plot.title = element_text(size = 9, vjust = -1, hjust = .6),
    axis.title.x=element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks.x=element_blank(), 
    axis.title.y = element_blank(), 
    axis.text.y = element_text(size = 8, margin = margin(r = -2)),
    axis.ticks.y=element_blank()) +
    ggtitle(paste0(down_pathways_levels[[4]]))
down_GO_3_lcpm_heat
```
##### down 4
```{r}
# lock in gene order which is sorted by adjusted p-value
GO_4_DEG <- subset(DEG_df, DEG_df$gene %in% GO_4_top$gene)
GO_4_DEG$gene <- factor(GO_4_DEG$gene,
                          levels = unique(GO_4_top$gene))
GO_4_DEG$gene <- fct_rev(GO_4_DEG$gene)

# heatmap 
down_GO_4_lcpm_heat <- ggplot(data = GO_4_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    name = expression(log[2](CPM))
  ) +
  scale_x_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(0.2, 0.5, 0.2, 0.2, "cm"), 
    panel.spacing = unit(0,'lines'), 
    plot.title = element_text(size = 9, vjust = -1),
    axis.title.x=element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks.x=element_blank(), 
    axis.title.y = element_blank(), 
    axis.text.y = element_text(size = 8, margin = margin(r = -2)),
    axis.ticks.y=element_blank()) +
    ggtitle(paste0(down_pathways_levels[[9]]))
down_GO_4_lcpm_heat
```
##### down 5
```{r}
# lock in gene order which is sorted by adjusted p-value
GO_5_DEG <- subset(DEG_df, DEG_df$gene %in% GO_5_top$gene)
GO_5_DEG$gene <- factor(GO_5_DEG$gene,
                          levels = unique(GO_5_top$gene))
GO_5_DEG$gene <- fct_rev(GO_5_DEG$gene)

# heatmap 
down_GO_5_lcpm_heat <- ggplot(data = GO_5_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    name = expression(log[2](CPM))
  ) +
  scale_x_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(0.2, 0.5, 0.2, 0.2, "cm"), 
    panel.spacing = unit(0,'lines'), 
    plot.title = element_text(size = 9, vjust = -1, hjust = .6),
    axis.title.x=element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks.x=element_blank(), 
    axis.title.y = element_blank(), 
    axis.text.y = element_text(size = 8, margin = margin(r = -2)),
    axis.ticks.y=element_blank()) +
    ggtitle(paste0(down_pathways_levels[[14]]))
down_GO_5_lcpm_heat
```

### legend only
```{r}
# Extract the legend. Returns a gtable
# lock in gene order which is sorted by adjusted p-value
GO_1_DEG <- subset(DEG_df, DEG_df$gene %in% GO_1_top$gene)
GO_1_DEG$gene <- factor(GO_1_DEG$gene,
                          levels = unique(GO_1_top$gene))
GO_1_DEG$gene <- fct_rev(GO_1_DEG$gene)

# heatmap 
legend_data <- ggplot(data = GO_1_DEG, aes(x = samples, y = gene)) +
  geom_tile(aes(fill = counts)) +
  ggplot2::facet_grid(~ Condition, scales = "free", switch = "both") +
  scale_fill_gradient2(
    low = "#FFFFCCFF",
    mid = "#FD8D3CFF",
    high = "#800026FF",
    midpoint = 6,
    space = "rgb",
    guide = "colourbar",
    breaks = c(0, 4, 8, 12, 16),
    limits = c(0,16),
    labels=c(0, 4, 8, 12, 16), 
    name = expression(log[2](CPM))
  ) +
  theme(
    legend.position = "top", 
    legend.text = element_text(size = 7.5), 
    legend.title = element_text (size =9), 
    legend.direction = "horizontal", 
    legend.key.size = unit(0.5, "cm"))
leg <- get_legend(legend_data)
# Convert to a ggplot and print
legend_tpm_bar <- as_ggplot(leg)
legend_tpm_bar

```
clean up
```{r, eval =FALSE}
remove(
  down_data,
  down_data_melt,
  down_df,
  down_enrich_df,
  down_enrich_df_melt,
  down_enrich_results_subset,
  down_gene_char,
  down_gene_pathway,
  down_input,
  GO_1,
  GO_1_DEG,
  GO_1_top,
  GO_2,
  GO_2_DEG,
  GO_2_top,
  GO_3,
  GO_3_DEG,
  GO_3_top,
  GO_4,
  GO_4_DEG,
  GO_4_top,
  GO_5, 
  GO_5_DEG,
  GO_5_top
)
```

# Combine all plots 
```{r}
row2 <-
  ggarrange(
    NULL,
    legend_tpm_bar,
    NULL,
    NULL,
    ncol = 4,
    labels = c("up-regulated", "", "", ""),
    font.label = list(size = 11, color = "red")
  )
row2

row3 <- ggarrange(
  up_GO_1_lcpm_heat,
  up_GO_2_lcpm_heat,
  up_GO_3_lcpm_heat,
  up_GO_4_lcpm_heat,
  up_GO_5_lcpm_heat,
  ncol = 5,
  labels = c("c)", "d)", "e)", "f)", "g)"),
  font.label = list(size = 11)
  )

row3

row4 <-
  ggarrange(
    NULL,
    NULL,
    NULL,
    NULL,
    ncol = 4,
    labels = c("down-regulated", "", "", ""),
    font.label = list(size = 11, color = "blue")
  )
row4

row5 <- ggarrange(
  down_GO_1_lcpm_heat,
  down_GO_2_lcpm_heat,
  down_GO_3_lcpm_heat,
  down_GO_4_lcpm_heat,
  down_GO_5_lcpm_heat,
  ncol = 5,
  labels = c("h)", "i)", "j)", "k)", "l)"),
  font.label = list(size = 11)
  )
row5

combind <-
  ggarrange(
    row1,
    row2,
    row3,
    row4,
    row5,
    nrow = 5,
    heights = c(4.5, 0.33, 3, 0.225, 3)
  )
combind
path <- paste0("../../../results/manuscript_figures/Figure_",
               tolower(tissue),
               "_v220701_5heatmaps_smaller_labels")
saveToPDF(paste0(path, ".pdf"), width = 10, height = 11)
```
# Session information
```{r}
session_info()
```
